---
title: "PBMC scRNA clustering using Poisson factorization"
author: "M Thomas"
date: "04/07/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 15, fig.height = 9)

library(Seurat)
library(patchwork)
library(poismf)
library(philentropy)
library(igraph)
library(dplyr)
library(tidyr)
library(ggplot2)
library(dbscan)
library(plotly)
library(RColorBrewer)
library(xtable)

set.seed(1234)
```

## Pre-Processing

The pre-processing is from the Seurat tutorial: https://satijalab.org/seurat/articles/pbmc3k_tutorial.html


```{r Pre-Processing}

pbmc.data <- readRDS('10X_PBMC.RData')
pbmc <- CreateSeuratObject(counts = pbmc.data, min.cells = 3, min.features = 200)

# QC
pbmc[['percent.mt']] <- PercentageFeatureSet(pbmc, pattern = '^MT-') # MT gene content
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5) # Remove low quality cells

# Normalize and scale for feature plots
pbmc <- NormalizeData(pbmc)
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)

```

## Compute Poisson factors

96 PFs are computed from counts for clustering, 30 for visualization

```{r Poisson Factors}

# Get counts from Seurat object
RNAdata <- pbmc@assays$RNA@counts
RNAdata <- SeuratDisk::Transpose(RNAdata)
RNAdata <- as.matrix(RNAdata)

# Calc. PFs
PF <- poismf(k=96, RNAdata, method="tncg")

colnames(PF$B) <- colnames(RNAdata)
colnames(PF$A) <- rownames(RNAdata)
df_A <- t(as.matrix(PF$A))
df_B <- t(as.matrix(PF$B))

# Quick plot of first two PFs
plot(df_A[,1], df_A[,2])

# Add reduction to Seurat object
pbmc[['hpf']] <- CreateDimReducObject(embeddings = df_A,
                                       loadings = df_B,
                                       key = 'pf_',
                                       assay = 'RNA')


# Repeat for visualization
PF_viz <- poismf(k=30, RNAdata, method="tncg")
colnames(PF_viz$B) <- colnames(RNAdata)
colnames(PF_viz$A) <- rownames(RNAdata)

df_A_viz <- t(as.matrix(PF_viz$A))
df_B_viz <- t(as.matrix(PF_viz$B))

pbmc[['pf_viz']] <- CreateDimReducObject(embeddings = df_A_viz,
                                       loadings = df_B_viz,
                                       key = 'pfviz_',
                                       assay = 'RNA')

```

## Clustering

Jensen-Shannon divergence is computed from counts giving a distance matrix. This is square rooted to produce the Jensen-Shannon distance.

An RNN graph is constructed using a JSD radius of 0.5. Smaller radii will produce smaller clusters, however, this also produces lots more unclustered points/ very small clusters.
The closest 25 points within the radius is kept

The JSD is used for the edge weights of the RNN graph. Louvain clustering is then used to cluster the resulting graph.

```{r Clustering}

# Calculate Jensen-Shannon Distance distance matrix
jsd <- sqrt(JSD(df_A, est.prob = 'empirical'))

# Histogram for selecting RNN radius
hist(jsd)

# Generate RNN graph
RNN <- frNN(jsd, 0.5, sort = T)$id
RNN_capped <- lapply(RNN, head, 25) # only include 25 closest points 
RNN_graph <- graph_from_adj_list(RNN_capped, mode = 'all', duplicate = F)

# Add JSD as weights to graph
edge_list <- as_edgelist(RNN_graph)
E(RNN_graph)$weight <- mapply(function(x, y){jsd[x, y]}, edge_list[,1], edge_list[,2])

# Louvain cluster RNN graph
louvain <- cluster_louvain(RNN_graph)

# Cluster counts
table(louvain$membership)

# Add cluster labels to Seurat object
pbmc@meta.data$RNN_clusters <- louvain$membership

```

## Visualization

UMAP for visualization using 30 PFs

As many of the clusters have very few elements, to reduce the number of clusters being visualized there are two plots. All clusters with n > 10 are plotted separately to those with n < 10. Most of n < 10 contains unclusterd points [See table(louvain$membership)]


```{r UMAP, fig.height=6, fig.width=9}
# UMAP
pbmc <- RunUMAP(pbmc, reduction = 'pf_viz', dims = 1:30)

# Plot clusters (n > 10)
pbmc_big <- subset(x = pbmc, subset = RNN_clusters %in% as.vector(which(table(pbmc@meta.data$RNN_clusters) > 10)))

cluster_colour_big <- colorRampPalette(brewer.pal(11, 'RdYlBu'))(length(unique(pbmc_big@meta.data$RNN_clusters)))
umap1 <- DimPlot(pbmc_big, reduction = "umap", group.by = 'RNN_clusters') +
  scale_color_manual(values = cluster_colour_big) + 
  theme_dark()
ggplotly(umap1)

# Plot small clusters (n <= 10); many have n = 1
pbmc_small <- subset(x = pbmc, subset = RNN_clusters %in% as.vector(which(table(pbmc@meta.data$RNN_clusters) <= 10)))

cluster_colour_small <- colorRampPalette(brewer.pal(11, 'RdYlBu'))(length(unique(pbmc_small@meta.data$RNN_clusters)))
umap2 <- DimPlot(pbmc_small, reduction = "umap", group.by = 'RNN_clusters') +
  scale_color_manual(values = cluster_colour_small) + 
  theme_dark()
ggplotly(umap2)
```

## Cluster validation

From the plots below some biological groupings can be associated to the clusters. Markers used are present in the tutorial web page. Some clusters are well defined (e.g. B cells), others are not (CD4+/CD8+ cells).


```{r Validation}

# Normalized marker gene plot
FeaturePlot(pbmc_big, features = c("IL7R", "CCR7", "CD14", "LYZ", "S100A4", 
                                            "MS4A1", "CD8A", "FCGR3A", "MS4A7", "GNLY", 
                                            "NKG7", "FCER1A", "CST3", "PPBP"
                                         ), reduction = 'umap')

# Normalized violin plot of markers
VlnPlot(pbmc_big, features = c("IL7R", "CCR7", "CD14", "LYZ", "S100A4", 
                           "MS4A1", "CD8A", "FCGR3A", "MS4A7", "GNLY", 
                           "NKG7", "FCER1A", "CST3", "PPBP"
                           ), log = TRUE, group.by = 'RNN_clusters')

# Table generator
Cells <- c('B', 'CD4+ T Naive', 'CD4+ Mem.', 'CD8T+', 'NK', 'DC', 'FCGR3A+ Mono', 'CD14+ Mono', 'Platlet')
Markers <- c('MS4A1', 'IL7R, CCR7', 'IL7R, S100A4', 'CD8A', 'GNLY, NKG7', 'FCER1A, CST3', 'FCGR3A, MS4A7', 'CD14, LYZ', 'PPBP')
Clusters <- c('2, 14', '6, 9, 12, 16', '1, 3, 13', '7, 10', '5', '17', '11', '4, 8, 20', 'individuals in pbmc_small')

tab <- data.frame(Cells=Cells, Markers=Markers, Clusters=Clusters)

tab

# Label UMAP; 21 is an unidentified cluster
Idents(pbmc_big) <- pbmc_big@meta.data$RNN_clusters
new.ids <- c("Naive CD4+", "B", "Mem CD4+", "CD14+ Mono", "NK", 
             "Mem CD4+", "CD8+ T", "CD14+ Mono", "Naive CD4+", "CD8+ T", 
             "FCGR3A+ Mono", "Naive CD4+", "Mem CD4+", "B", "Naive CD4+",
             "DC", "CD14+ Mono", "21")
names(new.ids) <- levels(pbmc_big)
pbmc_big <- RenameIdents(pbmc_big, new.ids)

DimPlot(pbmc_big, reduction = 'umap', label = T, pt.size = 2) + NoLegend()


```


## Session info

<details>
```{r}
  utils::sessionInfo()
```
</details>
